Legion: Expressing Locality and Independence with Logical Regions

### 2012-04-26 doc/sc2012/legion.tex
intro first pass

In this paper we describe Legion, a parallel programming system based
on using {\em logical regions} to describe the organization of data.
A logical region names a set of objects.  Logical regions are first
class values in Legion, and in particular may be passed as arguments
to functions that access the data in those regions, providing locality
information.  Logical regions may be {\em partitioned} into (usually)
disjoint {\em subregions}, providing information for determining independence of computations.  Furthermore,
computations access logical regions with particular {\em privileges} (e.g., {\em
  read-only} and {\em read-write}, among others) 
and {\em coherence} (e.g., {\em exclusive access} and {\em atomic access}, among others).

We begin with an extended example of a Legion program in Section~\ref{sec:ex}.
Each subsequent section describes a level in the Legion programming
system from the application down to the machine:

@@

\item We summarize Legion's static type system (Section~\ref{sec:type}), which serves two primary
  purposes.  First, pointers to regions are checked
  statically, eliminating the need for expensive runtime checks.
  Second, the invariant that a function $f$ accesses only
  regions passed as arguments to $f$, or subregions of those regions, is enforced.
  This property guarantees the Legion implementation can rely on
  preservation of independence---if two functions are passed disjoint
  region arguments, their computations access disjoint
data.

\item We describe the execution semantics for Legion and show that
analyzing region dependencies within a single function is sufficient to 
guarantee that a Legion execution preserves sequential program
behavior if regions are accessed with the strongest (exclusive)
coherence privilege (Section~\ref{sec:exec}).  This local scheduling property is central to 
the scalability of Legion on distributed memory machines.

\item We describe the design and implementation of the Legion {\em high-level runtime}
system (Section~\ref{sec:highlevel}), which performs scheduling of {\em tasks} (functions to be executed in parallel)
and implements each logical region as one or more {\em physical regions}.
The high-level runtime's scheduling algorithm is analogous to an out-of-order processor,
but works at the granularity of the tasks and regions in a function body rather than the instructions and registers
in a code block.  We describe the stages of the task execution pipeline, and how the high-level runtime coordinates parallel
scheduling decisions across the machine.  We also present the {\em mapping} interface, an API for incorporating
application-specific knowledge into the high-level runtime's decisions.

\item We describe the design and implementation of the Legion {\em low-level runtime} (Section~\ref{sec:lowlevel}),
a portability layer designed to abstract a wide variety of hardware, including multicore chips, clusters, and accelerators
such as GPUs.

\item We present the results of experiments on two applications running on a cluster of multicore processors with
attached GPUs (Section~\ref{sec:exp}).  The applications are both irregular and illustrate the capability of Legion to 
exploit locality and independence on such platforms.
Running on a 4 node cluster with 8 total GPUs 
and 4 levels of memory hierarchy, our implementation of Legion 
achieves up to 5.9X speedup over a single CPU-GPU node on these applications.

>>>>

In this paper we describe Legion, a parallel programming system based
on using {\em logical regions} to describe the organization of data.
A logical region names a set of objects.  Logical regions are first
class values in Legion, and in particular may be passed as arguments
to distinguished functions called {\em tasks} that access the data in those regions, providing locality
information.  Logical regions may be {\em partitioned} into (usually)
disjoint {\em subregions}, providing information for determining independence of computations.  Furthermore,
computations access logical regions with particular {\em privileges} (e.g., {\em
  read-only} and {\em read-write}, among others) 
and {\em coherence} (e.g., {\em exclusive access} and {\em atomic access}, among others).
Priveleges express how a task may use its region arguments, providing data dependence information
that is used to guide the extraction of parallelism (e.g., if two tasks access the same region with
{\em read-only} priveleges the two tasks can potentially be run in parallel), and coherence annotations
express the required semantics of concurrent accesses to the region.  For example, if tasks $x$ and $y$ access region $r$
with {\em exclusive} access, then the effects of $x$ and $y$ on $r$ are guaranteed to be the same as if they had executed
in the sequential order expressed by the program; on the other hand, if the tasks access $r$ with {\em atomic} coherence, then 
Legion guarantees only  atomicity of the tasks with respect to $r$: either $x$ appears to run entirely before $y$ or vice versa.

To introduce the programming model, we begin with an extended example of a circuit simulation programmed in Legion
in Section~\ref{sec:ex}. This example illustrates regions, tasks, permissions/coherence properties, the interactions
between them, and how these building blocks are assembled into a realistic Legion program.   Subsequent sections each
describe a contribution in the implementation and evaluation of Legion:

@@

\item We define a {\em software out-of order processor}, or SOOP, for scheduling tasks with region arguments in a manner
analogous to how out-of-order hardware schedulers process instructions with register arguments.  In addition to pipelining
the scheduling of tasks over several stages, our SOOP is also hierarchical and can extract nested parallelism (because tasks may recursively spawn subtasks) as well as concurrent and distributed across the machine (Section~\ref{sec:soop}).

\item A crucial part of orchestrating a program's execution is how
tasks are assigned (or {\em mapped}) to processors and how regions are
mapped to specific memory units within the machine.  Often there is
application-specific information that, if exploited, results in better
mappings than a generic, one-size-fits-all mapping strategy.  We describe
a {\em mapping interface} that allows programmers to give the SOOP a specification
of how to map tasks and regions.  This mapping API is carefully defined so that
any use-defined mapping strategy can only affect the performance of applications,
not their correctness.  Thus, user mappings only affect the time and resources required
to run the program and cannot change the program's output (Section~\ref{sec:mapping}).

\item We present the results of experiments on four applications: fluid-flow on a regular grid,
a three-level AMR code solving a simple heat transfer equation, a circuit simulation,
and LINPACK.  We run each of these four codes and best available reference versions on three different
clusters of multicore processors with attached GPUs: a small cluster of our own, a medium-sized cluster designed for visualization applications, and up to half of the Keeneland supercomputer (Section~\ref{sec:results}).
